Working from home and digital divides: resilience during the pandemic. Supplemental Material.

### 2021-08-20 presentations/RinR2021.Rmd
online appendix

<br>
Budnitz, H., and E. Tranos (2021) Working from home and digital divides: resilience during the pandemic. *Annals of the AAG*. In press. [link](https://doi.org/10.1080/24694452.2021.1939647)
<br>
<br>
[etranos.info/post/rir2021](https://etranos.info/post/rir2021)

@@

## WFH / Telecommuting
- WFM = economic resilience during Covid-19
- capability to WFM:
  - occupations
  - quality digital infrastructure
- Spatial clustering and dependence

## Digital divides
- 1st level: availability and quality of internet connectivity
- 2nd level: skills to effectively utilise digital technologies
- 3rd level: returns of internet usage
- Cross-level interactions

## Our approach
- Intersection of digital and material divides
- Economic resilience of places as digital technologies became an essential
tool of productivity during the pandemic
<br>
<br>
**internet services + occupations $\Rightarrow$ WFM $\Rightarrow$ economic resilience**

@@

## Experienced Broadband Speeds 
- An overall trend of increased testing from March to April and then a slight reduction from April to May
- $2019$ evening peak of testing, leisure activities, download speeds
- $2020$ **new morning peak**

@@

## Aggregation 
- Relatively scarce data to analyse as stream
- Composite week time series
- Mon-Friday, hourly $6:00$ -- $24:00$, excl. bank holidays
- $18$ hourly data points $x$ $5$ weekdays ($90$) for each local authority ($382$)
- Focus on upload speeds

## Time series clustering
- *Aim*: create clusters of local authorities with similar temporal signatures of internet speed
- Shape-based approach
- Match two separate time-series objects based on the similarity of their shapes 
- Distances between the shapes
- Simplicity: 
  - *k*-means
  - medoids, time-series objects with a minimal distance to all other cluster objects

@@

# stargazer(aux.up,
#           type = "text",
#           nobs = T,
#           digits = 3,
#           font.size = "footnotesize",
#           #float.env = "sidewaystable",
#           dep.var.caption = "",
#           title = "Auxiliary multinomial regression of upload speed clusters on socio-economic and geographic LAD variables",
#           header = F,
#           covariate.labels=c(#"pop. 16-64, 2018",
#                              "pop, 2018",
#                              #"pop. densuty, 2018",
#                              "job density, 2018",
#                              "distance to nearest met. area",
#                              "distance to London",
#                              "South of the UK",
#                              "managerial jobs, 2020",
#                              "tech jobs, 2020",
#                              "skilled trade jobs, 2020",
#                              "professional jobs, 2020",
#                              "administrative jobs, 2020",
#                              "service jobs, 2020",
#                              "machine operation jobs, 2020",
#                              "earnings, 2019",
#                              "n. business est. per hab., 2019",
#                              "furloughed per hab., 2020",
#                              "AM tests per hab., 2020",
#                              "Virgin Media \\%, 2020"),
#           add.lines = list(c("McFadden's R squared",
#                              round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
#                              round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
#                              round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
#                              , round(r2.up[1],3), round(r2.up[1],3)),
#                            (c("N",
#                             n.obs, n.obs, n.obs,
#                             n.obs, n.obs, n.obs,
#                             n.obs, n.obs, n.obs
#                             , n.obs, n.obs))))
```
## Clusters 6 and 9

- LADs in Cluster 6 are more likely to hold managerial, tech and professional jobs
- Least likely to be furloughed
- Slow and less reliable speeds
- High job density

@@

- Although speeds were slow and not as reliable during the morning peak as in cluster 9, the skills that enabled telecommuting
also enabled **greater returns** from doing so ($\Downarrow$ furloughed)
- LADs in cluster 9 were able to benefit both from reliable internet connections and populations
able to work from home to capitalise on their digital infrastructure
- But **lower returns** in the pandemic $\Rightarrow$ greater numbers furloughed

@@

- Clusters 3, 7 and 9 are on the right side of the 1st layer of digital divide
- Internet resilience supports, but also frustrates a wide range of other urban economies: 
  - $\Uparrow$ morning peak tests per capita run in cluster 6
- Being on the right side of the 2nd level digital divide had a **greater impact on economic resilience** (3rd level digital divide), than having quality internet connectivity 
  - $\Rightarrow$ cluster 1 and 6

@@

## References {.allowframebreaks}

>>>>

<br>
Budnitz, H., and E. Tranos (2021) Working from home and digital divides: resilience during the pandemic. *Annals of the AAG*. In press. [link](https://doi.org/10.1080/24694452.2021.1939647)
<br>
<br>
[etranos.info/post/rinr2021](https://etranos.info/post/rinr2021)

@@

## WFH / Telecommuting
- WFH = economic resilience during Covid-19

- Capability to WFH:
  - occupations
  - quality digital infrastructure

- Spatial clustering and dependence

## Digital divides
- 1st level: availability and quality of internet connectivity

- 2nd level: skills to effectively utilise digital technologies

- 3rd level: returns of internet usage

- Cross-level interactions

## Our approach
- Intersection of digital and material divides

- Economic resilience of places as digital technologies became an essential
tool of productivity during the pandemic
<br>
<br>
**internet services + occupations $\Rightarrow$ WFM $\Rightarrow$ economic resilience**

@@

## Experienced Broadband Speeds 
- An overall trend of increased testing from March to April and then a slight reduction from April to May in 2020, compared to less testing as Spring progresses in 2019

- $2019$ evening peak of testing, leisure activities, download speeds
- $2020$ **new morning peak** $\Rightarrow$ flatter hourly temporal profile

@@

## Aggregation 
- Relatively scarce data to analyse as stream

- Composite week time series

- Mon-Friday, hourly $6:00$ -- $24:00$, excl. bank holidays

- $18$ hourly data points $x$ $5$ weekdays ($90$) for each local authority ($382$)

- Focus on upload speeds

## Time series clustering
- *Aim*: create clusters of local authorities with similar temporal signatures of internet speed

- Shape-based approach
- Match two separate time-series objects based on the similarity of their shapes 
- Distances between the shapes
- Simplicity: 
  - *k*-means
  - medoids, time-series objects with a minimal distance to all other cluster objects

@@

# stargazer(aux.up,
#           type = "text",
#           nobs = T,
#           digits = 3,
#           font.size = "footnotesize",
#           #float.env = "sidewaystable",
#           dep.var.caption = "",
#           title = "Auxiliary multinomial regression of upload speed clusters on socio-economic and geographic LAD variables",
#           header = F,
#           covariate.labels=c(#"pop. 16-64, 2018",
#                              "pop, 2018",
#                              #"pop. densuty, 2018",
#                              "job density, 2018",
#                              "distance to nearest met. area",
#                              "distance to London",
#                              "South of the UK",
#                              "managerial jobs, 2020",
#                              "tech jobs, 2020",
#                              "skilled trade jobs, 2020",
#                              "professional jobs, 2020",
#                              "administrative jobs, 2020",
#                              "service jobs, 2020",
#                              "machine operation jobs, 2020",
#                              "earnings, 2019",
#                              "n. business est. per hab., 2019",
#                              "furloughed per hab., 2020",
#                              "AM tests per hab., 2020",
#                              "Virgin Media \\%, 2020"),
#           add.lines = list(c("McFadden's R squared",
#                              round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
#                              round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
#                              round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
#                              , round(r2.up[1],3), round(r2.up[1],3)),
#                            (c("N",
#                             n.obs, n.obs, n.obs,
#                             n.obs, n.obs, n.obs,
#                             n.obs, n.obs, n.obs
#                             , n.obs, n.obs))))
```
## Cluster 6

- Workers are more likely to hold managerial, tech and professional jobs
- Least likely to be furloughed
- Slow and less reliable speeds
- High job density

@@

- Although speeds were slow and not as reliable during the morning peak as in cluster 9, the skills that enabled telecommuting
also enabled **greater returns** from doing so ($\Downarrow$ furloughed)

- LADs in cluster 9 were able to benefit both from reliable internet connections and populations
able to work from home to capitalise on their digital infrastructure

- But **lower returns** in the pandemic $\Rightarrow$ greater numbers furloughed

@@

- Clusters 3, 7 and 9 are on the right side of the 1st layer of digital divide

- Internet resilience supports, but also frustrates urban economies in different geographies: 
  - $\Uparrow$ morning peak tests per capita run in cluster 6

- Being on the right side of the 2nd level digital divide had a **greater impact on economic resilience** (3rd level digital divide), than having quality internet connectivity 
  - $\Rightarrow$ cluster 1 and 6

@@

# References {.allowframebreaks}

### 2021-08-20 presentations/RinR2021_HB.Rmd
online appendix

<br>
Budnitz, H., and E. Tranos (2021) Working from home and digital divides: resilience during the pandemic. *Annals of the AAG*. In press. [link](https://doi.org/10.1080/24694452.2021.1939647)
<br>
<br>
[etranos.info/post/rinr2021](https://etranos.info/post/rinr2021)

@@

## WFH / Telecommuting
- From a niche means of accessing work to a widespread way of life
- From $5$% to $47$% in the UK [@ons2020; @ons2020lm2019]
- $37$% WFM in European April $2020$ 
- $60$% in Finland  [@eurofound2020]
- Almost $50$% in the US, $37$% could be permanently moved online [@brynjolfsson2020covid; @NBERw26948]

@@

- Economic resilience of places as digital technologies became an essential
tool of productivity during the pandemic
<br>
<br>
**internet services + occupations $\Rightarrow$ WFM $\Rightarrow$ economic resilience**

@@

- Composite week time series

@@

stargazer(clustersLADup[1:7],
          summary = F,
          font.size = "footnotesize",
          type = "html",
          header = F,
          digit.separator = ",",
          digit.separate = c(3,3),
          covariate.labels=c("Cluster",
                             "N. of LADs",
                             "LAD population",
                             "mean speed",
                             "SD speed",
                             "mean AM speed",
                             "mean PM speed"),
                             #"AM_slower",
                             #"SDtoMean"),
          rownames = F,
          no.space=TRUE,
          column.sep.width = "0pt",
          notes = "Note: All speed measures are upload speeds",
          title = "Upload speed cluster characteristics")
```

@@

- Workers are more likely to hold managerial, tech and professional jobs
- Least likely to be furloughed
- Slow and less reliable speeds
- High job density

@@

- Although speeds were slow and not as reliable during the morning peak as in cluster 9, the skills that enabled telecommuting
also enabled **greater returns** from doing so ($\Downarrow$ furloughed)

- LADs in cluster 9 were able to benefit both from reliable internet connections and populations
able to work from home to capitalise on their digital infrastructure

@@

- Clusters 3, 7 and 9 are on the right side of the 1st layer of digital divide

@@

- Being on the right side of the 2nd level digital divide had a **greater impact on economic resilience** (3rd level digital divide), than having quality internet connectivity 
  - $\Rightarrow$ cluster 1 and 6

@@

# References {.allowframebreaks}

>>>>

<br>
Budnitz, H., and E. Tranos (2021) Working from home and digital divides: resilience during the pandemic. *Annals of the AAG*. In press. [link](https://doi.org/10.1080/24694452.2021.1939647)
<br>
<br>
[etranos.info/post/rir2021](https://etranos.info/post/rir2021)

@@

## WFH / Telecommuting
- From a niche means of accessing work to a widespread way of life
- From $5$% to $47$% in the UK [@ons2020; @ons2020lm2019]
- $37$% WFH in Europe April $2020$ 
- Almost $50$% in the US, $37$% could be permanently moved online [@brynjolfsson2020covid; @NBERw26948]

@@

- Economic resilience of places as digital technologies became an essential
tool of productivity during the pandemic
<br>
<br>
**internet services + occupations $\Rightarrow$ WFH $\Rightarrow$ economic resilience**

@@

- Composite week time series for March-May 2020

@@

stargazer(clustersLADup[1:7],
          summary = F,
          font.size = "footnotesize",
          type = "html",
          header = F,
          digit.separator = ",",
          digit.separate = c(3,3),
          covariate.labels=c("Cluster",
                             "N. of LADs",
                             "LAD population",
                             "mean speed",
                             "SD speed",
                             "mean AM speed",
                             "mean PM speed"),
                             #"AM_slower",
                             #"SDtoMean"),
          rownames = F,
          no.space=TRUE,
          column.sep.width = "0pt",
          notes = "Note: All speed measures are upload speeds",
          title = "Upload speed cluster characteristics Spring 2020")
```

@@

- Workers in LADs more likely to hold managerial, tech and professional jobs
- Least likely to be furloughed
- Slow and less reliable speeds
- High job density

@@

- Although speeds in cluster 6 were slow and not as reliable during the morning peak as in cluster 9, the skills that enabled telecommuting
also enabled **greater returns** from doing so ($\Downarrow$ furloughed)

- LADs in cluster 9 were able to benefit from reliable internet connections and still included substantial populations
able to work from home to capitalise on their digital infrastructure

@@

- Some clusters, such as 9, are on the right side of the 1st layer of digital divide

@@

- Being on the right side of the 2nd level digital divide had a **greater impact on economic resilience** (3rd level digital divide), than having quality internet connectivity

@@

## References {.allowframebreaks}

### 2021-08-20 src/online_appendix.Rmd
online appendix

---
title: "Online Appendix"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    social: menu
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../paper/v2_taylor_francis/")
  })
---

```{r, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
library(rmarkdown)
library(tidyverse)
library(plotly)
library(lubridate)
library(gridExtra)
library(openair)
library(rgdal)
library(maps)
library(rprojroot)
library(kableExtra)
#library(cowplot)
library(ggthemes)

@@

# save clusters file
path.clusters <- paste(path, "/data/temp/clusters_nodiff_9.csv", sep = "")
clusters.nodiff <- read_csv(path.clusters)
```

### LAD, Upload speed clusters, 2020

```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE, fig.height=10, fig.asp=T}
#out.width='100%'

# LAD fully generalised
la <- readOGR("https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.geojson",
              verbose = FALSE)
# saved locally in /data/raw/Local_Authority_Districts_(December_2019)_Boundaries_UK_BUC.geojson

# spatial transformations
la <- spTransform(la, CRS("+init=epsg:4326"))
#la@data$LAD19NM <- as.character(la@data$LAD19NM)
la@data$lad19nm <- as.character(la@data$lad19nm)

@@

# fortify
la.f <- broom::tidy(la, region = "lad19nm")
la.f <- merge(la.f, clusters.nodiff, by.x = "id", by.y = "LAD")

# create the text to tooltip
la.f <- merge(la.f, for.tip.mean.up.am, by.x = "id", by.y = "LAD19NM", all.x=T)
la.f <- merge(la.f, for.tip.mean.up.pm, by.x = "id", by.y = "LAD19NM", all.x=T)

#sapply(la.f, function(x) sum(is.na(x)))
#No am data points for Isles of Scilly
la.f$tip <- paste0(la.f$id, "<br>",
                   "Mean upload AM speed: ", round(la.f$mean.up.am), " Kbps", "<br>",
                   "Mean upload PM speed: ", round(la.f$mean.up.pm), " Kbps")
la.f <- la.f[order(la.f$order),] # if i don't order merge.nuts.f loses order and the map has gaps

# using world.cities from `maps`
cities <- world.cities %>% filter(country.etc=="UK")

p <- ggplot(la.f, aes(x = long, y = lat)) +
  geom_polygon(aes(group = group, fill = as.factor(cluster.up), 
                   text = tip)) +# ,text =paste("test1", long, "test2", lat)
  #labs(title = "Upload clusters") +
  scale_fill_brewer(palette="Set1", 
                    direction = -1) +
  #geom_text_repel( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=4) +
  geom_point( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat), color="black", size=0.5) +
  geom_text( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=3) +

  guides(fill=guide_legend(title="Cluster")) +
  theme_map()
#   theme(panel.border = element_blank(), panel.grid.major = element_blank(),
# panel.grid.minor = element_blank())

p %>%
  ggplotly(tooltip = "text") %>%
  style(hoverlabel = list(bgcolor = "white"), hoveron = "fill")
```

@@

### LAD mean AM and PM upload speeds

>>>>

---
title: 'Working from home and digital divides: resilience during the pandemic. Supplemental Material.'
output: 
  flexdashboard::flex_dashboard:
    vertical_layout: fill
    orientation: columns
    social: menu
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../paper/v2_taylor_francis/")
  })
---

```{r, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
library(rmarkdown)
library(tidyverse)
library(lubridate)
library(gridExtra)
library(openair)
library(rgdal)
library(maps)
library(rprojroot)
library(kableExtra)
#library(cowplot)
library(ggthemes)
library(leaflet)

@@

# read clusters file
path.clusters <- paste(path, "/data/temp/clusters_nodiff_9.csv", sep = "")
clusters.nodiff <- read_csv(path.clusters)
```

Column {data-width=800}
-----------------------------------------------------------------------

### Upload speed clusters for Local Authority Districts, 2020. See [Budnitz and Tranos (2021)](https://doi.org/10.1080/24694452.2021.1939647) for more details.

```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE, fig.asp=T}

# LAD fully generalised
# la <- readOGR("https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.geojson",
#               verbose = FALSE)
# saved locally in /data/raw/Local_Authority_Districts_(December_2019)_Boundaries_UK_BUC.geojson

path.geojson <- paste0(path, "/data/raw/Local_Authority_Districts_(December_2019)_Boundaries_UK_BGC.geojson")
la <- readOGR(path.geojson, verbose = FALSE)

# spatial transformations
la <- spTransform(la, CRS("+init=epsg:4326"))

la@data$lad19nm <- as.character(la@data$lad19nm)

@@

# merge with the data
la <- sp::merge(la, clusters.nodiff, by.x = "lad19nm", by.y = "LAD")

# create the text to tooltip
la <- sp::merge(la, for.tip.mean.up.am, by.x = "lad19nm", by.y = "LAD19NM", all.x=T)
la <- sp::merge(la, for.tip.mean.up.pm, by.x = "lad19nm", by.y = "LAD19NM", all.x=T)

#sapply(la@data, function(x) sum(is.na(x)))
#No am data points for Isles of Scilly

# text for labels
la@data$tip <- paste0(la@data$lad19nm, "<br>",
                   "Mean upload AM speed: ", round(la@data$mean.up.am), " Kbps", "<br>",
                   "Mean upload PM speed: ", round(la@data$mean.up.pm), " Kbps") %>% 
  lapply(htmltools::HTML)

# using world.cities from `maps`
cities <- world.cities %>% 
  filter(country.etc=="UK") %>% 
  arrange(pop) %>% tail(10)

# Create a color palette for the map:
mypalette <- colorFactor(palette="Set1", 
                         domain=la$cluster.up, 
                         na.color="transparent",
                         reverse = T)

leaflet(la) %>% 
  addTiles()  %>% 
  fitBounds(-5.7, 50, -1.2, 58.7) %>% 
  addPolygons(fillColor = ~mypalette(cluster.up), 
              stroke=T,
              color = "white",
              weight = .8,
              opacity = .3,
              fillOpacity = .5,
              label = ~tip) %>% 
  addCircles(lng = cities$long, lat = cities$lat,
             weight = 2, radius = 5, color = "black",
             label = cities$name,
             labelOptions = labelOptions(noHide = T, direction = 'right', 
                                         textOnly = TRUE, permanent = F),
             options = markerOptions(riseOnHover = TRUE)) %>% 
  # addLabelOnlyMarkers(data = cities,
  #                   lng = ~long, lat = ~lat, label = ~name,
  #                   labelOptions = labelOptions(noHide = T, direction = 'auto', textOnly = TRUE)) %>%
  addProviderTiles(providers$Esri.WorldTerrain) %>% 
  addLegend(pal = mypalette,
            values = ~cluster.up,
            position = "topleft",
            opacity = .5,
            title = "Upload speed<br>clusters")

```

@@

### Local Authority District mean AM and PM upload speeds
